---
title: "Week 12: Neural Networks — Layers, Signals, and Learning from Examples"
author: "Roozbeh Sadeghian"
date: "Last Updated: `r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    highlight: tango
    code_folding: show
    df_print: paged
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@2/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
---

```{r setup, include=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 6
)

if (!requireNamespace("seedhash", quietly = TRUE)) {
  if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes", repos = "https://cloud.r-project.org")
  }
  remotes::install_github("roozbeh412/seedhash", subdir = "R", upgrade = "never")
}
```

```{r seeds, include=TRUE}
library(seedhash)
gen <- SeedHashGenerator$new("ANLY530-Week12-NeuralNetworks")
MASTER_SEED <- gen$generate_seeds(1)
set.seed(MASTER_SEED)
```

```{r seedhash-info, echo=FALSE, results='asis'}
cat(sprintf(
  "**Random Seed Source:** `seedhash` package — input string: `\"ANLY530-Week12-NeuralNetworks\"` → seed: `%d`\n\n",
  MASTER_SEED
))
```

# Introduction

This week we study **neural networks**: flexible models built from many tiny calculators (**neurons**) wired together. The wiring strengths (**weights**) and small offsets (**biases**) are learned from data—much like you adjusted knobs when tuning a radio until the station came in clearly.

> **Plain English:** A neural network is a **recipe machine**. Raw ingredients (your features) go in one side. Each neuron mixes its inputs with learned weights, adds a bias, and passes the result through an **activation function**. The final layer produces a prediction (a class label, a probability, or a number). Training repeatedly adjusts weights so predictions move closer to the correct answers.

Neural networks appear under buzzwords like **deep learning** or **AI**, but the core ideas match what you already know from regression and logistic regression: combine inputs linearly, then apply a nonlinear transform. The difference is **depth** and **width**: many chained transformations let the model represent rich, curved boundaries instead of a single straight line or plane.

These beginner-friendly articles explain the same story with worked examples: GeeksforGeeks [1] and Victor Zhou’s feedforward walk-through [2]. Your lecture slides (`ANLY530Lecture13_NeuralNetworks.pptx`) anchor the course terminology.

**Learning objectives**

1. Name the parts of a network (**input / hidden / output** layers, **weights**, **biases**, **activations**).
2. Trace **forward propagation** with a small numeric example.
3. Interpret common **activation functions** and why **nonlinearity** matters (the XOR demo).
4. Connect **loss**, **gradients**, **backpropagation**, and **gradient descent** with pictures—noticing how **learning rate** changes the path.
5. Train a simple **feedforward** classifier in R (`nnet::nnet`) and compare it to **multinomial logistic regression** (`nnet::multinom`).
6. Discuss **capacity**, **overfitting**, and the **black-box** trade-off honestly.

## Required packages

```{r load-packages}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  dplyr,
  tibble,
  purrr,
  ggplot2,
  nnet,
  scales,
  knitr,
  RColorBrewer
)
```

---

# Part 1: Vocabulary in plain English

| Term | Plain English |
|:-----|:--------------|
| **Neuron** | One computational unit: weighted sum of inputs + bias → activation. |
| **Layer** | A row of neurons at the same depth in the network. |
| **Input layer** | Holds your measured features (often shown as nodes for bookkeeping—not always “computed”). |
| **Hidden layer** | Layers between input and output; they build intermediate representations. |
| **Output layer** | Produces the prediction (e.g., class probabilities or a numeric value). |
| **Weight** | How strongly an input influences the next neuron. Learned from data. |
| **Bias** | Lets the neuron shift its activation up or down even when inputs are zero. Also learned. |
| **Activation function** | Nonlinear squeeze/split (sigmoid, ReLU, …) that lets the network bend boundaries. |
| **Forward pass** | Data flows input → hidden → output with current weights. |
| **Loss** | A score measuring how wrong predictions are; training tries to lower it. |
| **Backpropagation** | The recipe that computes how each weight affected the loss (uses the chain rule). |
| **Epoch** | One full sweep through the training data (possibly in mini-batches). |
| **Learning rate** | Step size when updating weights; too large oscillates, too small crawls. |

---

# Part 2: One neuron = logistic regression’s cousin

A single neuron computes

$$z = w_1 x_1 + w_2 x_2 + \cdots + w_p x_p + b, \qquad a = f(z)$$

where $f$ is the activation. If $f$ is the **sigmoid** $\sigma(z) = 1/(1+e^{-z})$, one neuron is exactly **logistic regression** logic: a weighted sum followed by a probability-friendly squash into $(0,1)$.

```{r single-neuron-example}
# Victor Zhou’s toy numbers: w = [0, 1], b = 4, x = [2, 3]
w <- c(0, 1)
b <- 4
x <- c(2, 3)
z <- sum(w * x) + b
sigmoid <- function(z) 1 / (1 + exp(-z))
a <- sigmoid(z)
cat(sprintf("Weighted sum z = %.4f; sigmoid(z) = %.4f\n", z, a))
```

```{r activation-comparison, fig.height=5, fig.width=10}
zseq <- seq(-6, 6, length.out = 400)
nz <- length(zseq)
act_df <- tibble(
  zgrid = rep(zseq, 3),
  activation = factor(
    rep(c("sigmoid", "tanh", "relu"), each = nz),
    levels = c("sigmoid", "tanh", "relu")
  ),
  value = c(1 / (1 + exp(-zseq)), tanh(zseq), pmax(0, zseq))
)

ggplot(act_df, aes(x = zgrid, y = value, color = activation)) +
  geom_line(size = 1.1) +
  geom_hline(yintercept = 0, color = "grey70") +
  geom_vline(xintercept = 0, color = "grey70") +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Three Common Activation Functions",
    subtitle = "Sigmoid and tanh squash values; ReLU clips negatives to zero (sparsity + simpler gradients)",
    x = "Input z (before activation)",
    y = "Output f(z)",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

> **Reading the plot:** **Sigmoid** outputs probabilities in $(0,1)$—nice for single-output classification. **Tanh** centers outputs around zero. **ReLU** ($\max(0,z)$) is piecewise linear; it trains efficiently and is the default in many deep frameworks [1].

---

# Part 3: Wiring neurons into a **feedforward** network

Data moves one way: input → hidden(s) → output. Below is a schematic **2 → 2 → 1** network (two inputs, two hidden neurons, one output)—the same shape Victor Zhou uses to illustrate feedforward [2].

```{r network-diagram, fig.height=4.8, fig.width = 9, echo=FALSE}
layers <- list(
  input = tibble(id = c("x1", "x2"), x = c(1, 1), y = c(3, 1)),
  hidden = tibble(id = c("h1", "h2"), x = c(3, 3), y = c(3.4, 0.6)),
  output = tibble(id = "y_hat", x = 5, y = 2)
)

edges <- bind_rows(
  as_tibble(merge(data.frame(from = layers$input$id), data.frame(to = layers$hidden$id))) %>%
    mutate(layer = "in_hidden"),
  tibble(from = layers$hidden$id, to = "y_hat", layer = "hidden_out")
)

node_xy <- bind_rows(
  layers$input %>% mutate(layer = "Input"),
  layers$hidden %>% mutate(layer = "Hidden"),
  layers$output %>% mutate(layer = "Output")
)

from_xy <- node_xy %>% rename(x0 = x, y0 = y, from = id)
to_xy <- node_xy %>% rename(x1 = x, y1 = y, to = id)

seg <- edges %>%
  left_join(from_xy %>% select(from, x0, y0), by = "from") %>%
  left_join(to_xy %>% select(to, x1, y1), by = "to")

ggplot() +
  geom_segment(
    data = seg,
    aes(x = x0, y = y0, xend = x1, yend = y1),
    size = 0.9,
    color = "grey55",
    arrow = arrow(length = unit(0.14, "inches"))
  ) +
  geom_label(
    data = node_xy,
    aes(x = x, y = y, label = id),
    size = 5,
    fontface = "bold",
    fill = "#1E88E5",
    color = "white",
    label.size = 0
  ) +
  annotate(
    "label",
    x = 2,
    y = 4.25,
    label = "Weights on every arrow\n(biases inside neurons)",
    fill = "#FFF3E0",
    size = 4
  ) +
  labs(
    title = "Feedforward Network (conceptual)",
    subtitle = "Hidden layer combines inputs; output neuron combines hidden activations"
  ) +
  theme_void(base_size = 13) +
  coord_equal(xlim = c(0.5, 5.6), ylim = c(0.2, 4.6), expand = FALSE)
```

> **Hidden layers** are not mysterious—they are **feature factories**. Early layers might detect edges or keyword flags; deeper layers combine those pieces. For tabular data (rows and columns), each hidden neuron still plays the same mathematical role: weighted sum + activation.

---

# Part 4: Numeric forward pass (spam-keyword style)

GeeksforGeeks walks through a tiny email-spam example [1]: binary keyword features → hidden ReLU → output sigmoid. Below is the same calculation in R so you can line up code with the article’s tables.

```{r forward-pass-spam-example}
x <- c(1, 0, 1) # presence of keywords: free, win, offer
W_h <- rbind(c(0.5, -0.2, 0.3), c(0.4, 0.1, -0.5))
b_h <- c(0, 0)
z_h <- as.vector(W_h %*% x + b_h)
h <- pmax(0, z_h)

w_o <- c(0.7, 0.2)
b_o <- 0
z_o <- sum(w_o * h) + b_o
y_hat <- sigmoid(z_o)

cat("Hidden pre-activations z_h:", paste(round(z_h, 4), collapse = ", "), "\n")
cat("Hidden activations ReLU(z_h):", paste(round(h, 4), collapse = ", "), "\n")
cat(sprintf("Output logit z_o = %.4f; sigmoid(z_o) = %.4f\n", z_o, y_hat))
cat(sprintf(
  "Decision at 0.5 threshold: %s\n",
  ifelse(y_hat >= 0.5, "Spam (1)", "Not spam (0)")
))
```

This is **forward propagation**: multiply-add, activate, repeat layer by layer.

---

# Part 5: Why nonlinearity matters — the XOR puzzle

With **only linear boundaries**, you cannot perfectly separate XOR: points $(0,0)$ and $(1,1)$ share a label, while $(0,1)$ and $(1,0)$ share the opposite label.

```{r xor-setup}
xor_df <- tibble(
  x1 = c(0, 0, 1, 1),
  x2 = c(0, 1, 0, 1),
  class = factor(c("A", "B", "B", "A"))
)
```

```{r xor-linear-vs-nnet, fig.height = 5.5, fig.width = 10}
grid <- as_tibble(expand.grid(
  x1 = seq(-0.2, 1.2, length.out = 120),
  x2 = seq(-0.2, 1.2, length.out = 120)
))

# Linear logistic regression boundary
glm_xor <- glm(as.numeric(class == "B") ~ x1 + x2, data = xor_df, family = binomial)
p_lin <- predict(glm_xor, grid, type = "response")
grid_lin <- grid %>% mutate(prob_B = p_lin)

set.seed(MASTER_SEED)
nn_xor <- nnet::nnet(class ~ x1 + x2, data = xor_df, size = 4, decay = 1e-4,
                     trace = FALSE, maxit = 500, linout = FALSE)
p_nn <- predict(nn_xor, grid, type = "raw")
if (is.matrix(p_nn)) {
  prob_B_nn <- p_nn[, "B"]
} else {
  prob_B_nn <- as.vector(p_nn)
}
grid_nn <- grid %>% mutate(prob_B = prob_B_nn)

panels <- c(
  "Logistic regression (no hidden layer)",
  "Feedforward net (hidden units)"
)
combined <- bind_rows(
  mutate(grid_lin, panel = panels[1]),
  mutate(grid_nn, panel = panels[2])
) %>%
  mutate(panel = factor(panel, levels = panels))

xor_pts <- bind_rows(
  mutate(xor_df, panel = panels[1]),
  mutate(xor_df, panel = panels[2])
) %>%
  mutate(panel = factor(panel, levels = panels))

ggplot(combined, aes(x = x1, y = x2, fill = prob_B)) +
  geom_raster(interpolate = TRUE) +
  facet_wrap(~panel, nrow = 1) +
  geom_point(
    data = xor_pts,
    aes(x = x1, y = x2, shape = class, color = class),
    size = 5,
    inherit.aes = FALSE,
    stroke = 1.2
  ) +
  geom_contour(aes(z = prob_B), breaks = 0.5, color = "black", size = 0.8) +
  scale_fill_viridis_c(option = "C", limits = c(0, 1), name = "P(class B)") +
  scale_color_manual(values = c(A = "#E65100", B = "#0277BD")) +
  scale_shape_manual(values = c(A = 16, B = 17)) +
  coord_fixed() +
  labs(
    title = "Same four XOR points — different model flexibility",
    subtitle = "Left: a single linear boundary; right: a bent boundary from hidden nonlinearities",
    x = expression(x[1]),
    y = expression(x[2])
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

> **Takeaway:** Stack neurons + nonlinear activations and boundaries become **curved**. Depth isn’t magic—it’s structured flexibility you must regularise so it doesn’t memorise noise [1].

---

# Part 6: Loss, gradients, and gradient descent (picture intuition)

Training minimizes a **loss** (mean squared error for regression; cross-entropy for classification). Think of loss as elevation on a hilly landscape where **weights are coordinates**. **Gradient descent** rolls downhill; **backpropagation** computes the downhill direction efficiently using calculus chain rules [1], [2].

```{r loss-surface-demo, fig.height = 5.5, fig.width = 9}
# Synthetic convex bowl L(w1,w2) — illustrative only
wgrid <- as_tibble(expand.grid(
  w1 = seq(-2.5, 2.5, length.out = 120),
  w2 = seq(-2.5, 2.5, length.out = 120)
)) %>%
  mutate(loss = 0.6 * w1^2 + 1.2 * w2^2 + 0.15 * sin(3 * w1) * cos(2 * w2))

# Gradient descent path from a random start
start <- c(2.1, -1.8)
eta <- 0.35
path <- vector("list", 22)
w <- start
for (i in seq_along(path)) {
  g1 <- 1.2 * w[1] + 0.45 * cos(3 * w[1]) * cos(2 * w[2])
  g2 <- 2.4 * w[2] - 0.3 * sin(3 * w[1]) * sin(2 * w[2])
  path[[i]] <- c(w1 = w[1], w2 = w[2])
  w <- w - eta * c(g1, g2)
}
path_df <- as_tibble(do.call(rbind, path)) %>% mutate(step = row_number())

ggplot(wgrid, aes(x = w1, y = w2, z = loss)) +
  geom_contour_filled(bins = 14, alpha = 0.85) +
  geom_path(data = path_df, aes(x = w1, y = w2, group = 1), inherit.aes = FALSE,
            color = "white", size = 1.2, arrow = arrow(length = unit(0.12, "inches"))) +
  geom_point(data = path_df, aes(x = w1, y = w2), inherit.aes = FALSE,
             color = "black", size = 2) +
  labs(
    title = "Illustrative Loss Landscape + Gradient Descent Path",
    subtitle = "White path: repeated small steps opposite to the gradient (learning rate η controls step size)",
    x = expression(w[1]), y = expression(w[2]), fill = "Loss"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

```{r learning-rate-compare, fig.height = 4.5, fig.width = 9}
quad_loss <- function(w) 0.5 * (w - 1.2)^2
quad_grad <- function(w) (w - 1.2)

run_gd <- function(eta, steps = 25, w0 = 3) {
  w <- w0
  out <- numeric(steps)
  for (i in seq_len(steps)) {
    out[i] <- w
    w <- w - eta * quad_grad(w)
  }
  tibble(step = seq_len(steps), w = out, loss = quad_loss(out), eta = factor(eta))
}

lr_curve <- bind_rows(
  run_gd(0.05),
  run_gd(0.55),
  run_gd(1.05)
)

ggplot(lr_curve, aes(x = step, y = loss, color = eta)) +
  geom_line(size = 1.1) +
  scale_y_log10() +
  labs(
    title = "Effect of Learning Rate on a Simple Bowl",
    subtitle = "Too small: slow; too large: overshoots and oscillates (loss spikes)",
    x = "Iteration",
    y = "Loss (log scale)",
    color = expression(eta)
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

# Part 7: Train a feedforward network on Iris (beginner-friendly benchmark)

We compare **multinomial logistic regression** (`nnet::multinom`) with a single-hidden-layer **feedforward network** (`nnet::nnet`). Iris is tiny—expect small gaps—but the pattern mirrors larger projects: hold out a test set, fit models, compare accuracy. Neural nets benefit from **scaled** numeric inputs, so we **center and scale** the four measurement columns using training-set means and SDs before fitting `nnet` (logistic regression is invariant to linear scaling of inputs, so we keep it on raw units for simplicity).

```{r iris-split}
data(iris)
set.seed(MASTER_SEED)
n_train <- floor(0.75 * nrow(iris))
train_idx <- sample(seq_len(nrow(iris)), size = n_train)
iris_train <- iris[train_idx, ]
iris_test <- iris[-train_idx, ]

scale_numeric_like_train <- function(train_df, test_df) {
  nums <- c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")
  train_s <- train_df
  test_s <- test_df
  for (v in nums) {
    mu <- mean(train_df[[v]])
    sdv <- stats::sd(train_df[[v]])
    if (!is.finite(sdv) || sdv < 1e-8) sdv <- 1
    train_s[[v]] <- (train_df[[v]] - mu) / sdv
    test_s[[v]] <- (test_df[[v]] - mu) / sdv
  }
  list(train = train_s, test = test_s)
}

iris_scaled <- scale_numeric_like_train(iris_train, iris_test)
iris_train_nn <- iris_scaled$train
iris_test_nn <- iris_scaled$test
```

```{r iris-models, results='hide'}
set.seed(MASTER_SEED)
log_fit <- nnet::multinom(Species ~ ., data = iris_train, trace = FALSE)

set.seed(MASTER_SEED)
nnet_fit <- nnet::nnet(
  Species ~ .,
  data = iris_train_nn,
  size = 10,
  decay = 0.01,
  maxit = 400,
  trace = FALSE,
  linout = FALSE,
  MaxNWts = 8000
)
```

```{r iris-results}
pred_log <- predict(log_fit, iris_test, type = "class")
pred_nn <- predict(nnet_fit, iris_test_nn, type = "class")

acc_log <- mean(pred_log == iris_test$Species)
acc_nn <- mean(pred_nn == iris_test$Species)

cat(sprintf("Hold-out accuracy — multinomial logistic: %.3f\n", acc_log))
cat(sprintf("Hold-out accuracy — nnet (10 hidden units, fixed decay): %.3f\n", acc_nn))
```

```{r iris-results-viz, fig.height = 4.5, fig.width = 7}
compare_df <- tibble(
  Model = c("Multinomial logistic", "Single-hidden-layer nnet"),
  Accuracy = c(acc_log, acc_nn)
)

ggplot(compare_df, aes(x = reorder(Model, Accuracy), y = Accuracy, fill = Model)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = sprintf("%.2f%%", Accuracy * 100)), hjust = -0.05, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(labels = percent, limits = c(0, 1.05)) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(
    title = "Iris Test Accuracy — Linear Softmax vs. One Hidden Layer",
    subtitle = "On easy separable data both can excel; networks shine when boundaries bend",
    x = NULL,
    y = "Accuracy"
  ) +
  theme_minimal(base_size = 12)
```

```{r iris-confusion, fig.height = 4.2, fig.width = 9}
cm_both <- bind_rows(
  mutate(as_tibble(table(Predicted = pred_log, Actual = iris_test$Species)), model = "Logistic regression"),
  mutate(as_tibble(table(Predicted = pred_nn, Actual = iris_test$Species)), model = "nnet")
) %>%
  mutate(model = factor(model, levels = c("Logistic regression", "nnet")))

ggplot(cm_both, aes(x = Actual, y = Predicted, fill = n)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = n), fontface = "bold", size = 6) +
  scale_fill_gradient(low = "white", high = "#3949AB") +
  facet_wrap(~model, nrow = 1) +
  labs(title = "Confusion matrices — Iris test set", x = "Actual", y = "Predicted") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

```{r iris-hidden-size-curve, fig.height = 4.5, fig.width = 8}
sizes_try <- c(2L, 4L, 8L, 12L, 16L)

acc_by_size <- map_dbl(sizes_try, function(sz) {
  set.seed(MASTER_SEED)
  fit_sz <- nnet::nnet(
    Species ~ .,
    data = iris_train_nn,
    size = sz,
    decay = 0.02,
    maxit = 350,
    trace = FALSE,
    linout = FALSE,
    MaxNWts = 12000
  )
  mean(predict(fit_sz, iris_test_nn, type = "class") == iris_test$Species)
})

curve_sz <- tibble(hidden_units = sizes_try, test_accuracy = acc_by_size)

ggplot(curve_sz, aes(x = hidden_units, y = test_accuracy)) +
  geom_line(size = 1.1, color = "#3949AB") +
  geom_point(size = 3, color = "#E65100") +
  scale_y_continuous(labels = scales::percent, limits = c(NA, 1)) +
  labs(
    title = "Iris: Test Accuracy vs. Number of Hidden Units",
    subtitle = "Quick sweep (same decay/maxit for each point)—caret would cross-validate this more rigorously",
    x = "Hidden units (size)",
    y = "Hold-out accuracy"
  ) +
  theme_minimal(base_size = 12)
```

---

# Part 8: Capacity, data hunger, and interpretability

```{r pros-cons-table, echo=FALSE}
trade <- tibble(
  Topic = c(
    "Flexibility",
    "Data needs",
    "Training cost",
    "Interpretability",
    "Overfitting risk"
  ),
  Guidance = c(
    "Hidden layers can bend boundaries; depth stacks bends.",
    "Large nets crave large, representative labeled sets [1].",
    "Many weights → longer training; GPUs help big models.",
    "Weights are hard to narrate—treat as high-accuracy black boxes unless using diagnostics.",
    "Match capacity to data; use validation curves, dropout (in deep libs), weight decay."
  )
)

knitr::kable(trade, caption = "Practical reminders for beginners")
```

---

# Knowledge check

<details>
<summary>**Question 1:** In one sentence, what does a hidden layer do?</summary>
**Answer:** It transforms the original features through learned weighted sums and nonlinear activations so the output layer can separate patterns that were not linearly separable in raw space.
</details>

<br>

<details>
<summary>**Question 2:** Why must activations be nonlinear?</summary>
**Answer:** Composing only linear layers stays linear overall—no new bending power. Nonlinear activations let stacked layers approximate curved boundaries (see XOR).
</details>

<br>

<details>
<summary>**Question 3:** Forward vs. backward pass?</summary>
**Answer:** Forward computes predictions with current weights; backward propagates loss derivatives to adjust each weight efficiently [1], [2].
</details>

<br>

<details>
<summary>**Question 4:** What is the learning rate?</summary>
**Answer:** It scales each gradient step; too large yields unstable training, too small wastes iterations (illustrated in Part 6).
</details>

<br>

<details>
<summary>**Question 5:** When might logistic regression be enough?</summary>
**Answer:** When classes are roughly linearly separable and interpretability matters more than squeezing the last points of accuracy—simple baselines first.
</details>

---

# Summary

1. Neural networks chain **weighted sums + biases + activations**; depth adds representational flexibility.
2. **Forward propagation** is multiply-add-activate layer by layer; **loss** scores mistakes.
3. **Backpropagation + gradient descent** adjust weights to reduce loss; **learning rate** matters.
4. Nonlinear activations unlock patterns like **XOR**; linear models cannot bend enough alone.
5. In R, `nnet`/`caret` trains compact feedforward models; match **capacity** to data and watch **generalisation**.

---

# References

[1] GeeksforGeeks, “Neural Networks — A beginner’s guide,” 2026. [Online]. Available: https://www.geeksforgeeks.org/deep-learning/neural-networks-a-beginners-guide/

[2] V. Zhou, “Machine Learning for Beginners: An Introduction to Neural Networks,” victorzhou.com, updated Jan. 13, 2026. [Online]. Available: https://victorzhou.com/blog/intro-to-neural-networks/

[3] Roozbeh Sadeghian, *ANLY530 Lecture 13 — Neural Networks* (PowerPoint slides), Week 12 course materials.

[4] Z. Huang, “seedhash: Deterministic Seed Generation from String Inputs Using MD5 Hashing,” R package, 2026. [Online]. Available: https://github.com/roozbeh412/seedhash
