---
title: "Week 06: Naive Bayes (NB) Classifiers — A Beginner-Friendly Guide"
author: "Ziyuan Huang"
date: "Last Updated: `r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    highlight: tango
    code_folding: show
    df_print: paged
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@2/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 6
)

# Deterministic seed via seedhash — reproducible across machines
if (!requireNamespace("seedhash", quietly = TRUE)) {
  if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages("pak", repos = "https://r-lib.github.io/p/pak/stable/")
  }
  pak::pkg_install("github::melhzy/seedhash/R", ask = FALSE)
}
```

```{r seeds, include=TRUE}
library(seedhash)
gen <- SeedHashGenerator$new("ANLY530-Week06-NaiveBayes")
MASTER_SEED <- gen$generate_seeds(1)
set.seed(MASTER_SEED)
```

```{r seedhash-info, echo=FALSE, results='asis'}
cat(sprintf("**Random Seed Source:** `seedhash` package — input string: `\"ANLY530-Week06-NaiveBayes\"` → seed: `%d`\n\n", MASTER_SEED))
```

# Introduction

This week we study **Naive Bayes (NB)** classifiers — one of the simplest, fastest, and most surprisingly effective classification methods.

If you are new to Machine Learning, NB is a great “first probability model” because:

- It forces you to think in **probabilities** (“How likely is spam?”) instead of just hard rules.
- The training step is mostly counting + simple averages (very fast).
- It is a classic baseline for **text classification** (spam filtering, sentiment).

We will follow Lecture 07 (`Lecture07-NB.pdf`) and two recommended references:

- Wikipedia: Naive Bayes classifier — `https://en.wikipedia.org/wiki/Naive_Bayes_classifier`
- scikit-learn user guide (conceptual overview): `https://scikit-learn.org/stable/modules/naive_bayes.html`

> **Important note:** We will implement NB in **R**, but the ideas match the Python examples in the slides and in scikit-learn.

## Learning Objectives

By the end of this tutorial, you will be able to:

1. Explain **prior**, **likelihood**, **posterior**, and **evidence** in plain English.
2. Compute a Bayes posterior probability by hand for a simple example.
3. Explain what the “**naive**” (independence) assumption is, and why it can still work.
4. Train and evaluate a **Gaussian Naive Bayes** model on the **Iris** dataset.
5. Train and evaluate a **Multinomial Naive Bayes** model on a tiny text dataset.
6. Understand **Laplace smoothing** and the “zero-frequency” problem.

## Required Packages

```{r load-packages}
if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  tidyverse,
  scales,
  gridExtra,
  broom,
  naivebayes,
  caret,
  pROC,
  tidytext
)
```

---

# Part 1: Bayes’ Theorem (The “Probability Update” Rule)

## 1.1 The Bayes Story in One Sentence

**Bayes’ Theorem** is a way to update your belief after seeing new evidence:

- Start with a **prior** belief: “How common is spam in general?”
- See evidence: “This email contains the word *Urgent*.”
- Update to a **posterior** belief: “Given *Urgent*, how likely is spam now?”

## 1.2 The Formula (and the Meaning of Each Piece)

Bayes’ Theorem:

$$P(A \mid B) = \frac{P(B \mid A)\,P(A)}{P(B)}$$

Where:

- \(P(A)\) is the **prior** probability of event \(A\)
- \(P(B \mid A)\) is the **likelihood** of evidence \(B\) if \(A\) is true
- \(P(B)\) is the **evidence** (also called marginal likelihood)
- \(P(A \mid B)\) is the **posterior** probability after seeing \(B\)

## 1.3 Example (from Lecture 07): “Urgent” Spam Email

From the slide story:

- \(P(\text{spam}) = 0.20\)  (20% of all emails are spam)
- \(P(\text{ham}) = 0.80\)
- \(P(\text{Urgent}) = 0.05\)  (5% of all emails contain “Urgent”)
- Suppose \(P(\text{Urgent} \mid \text{spam}) = 0.30\) (spam often uses urgent language)

Then:

$$P(\text{spam} \mid \text{Urgent}) = \frac{P(\text{Urgent} \mid \text{spam})\,P(\text{spam})}{P(\text{Urgent})}$$

```{r urgent-example}
P_spam <- 0.20
P_urgent <- 0.05
P_urgent_given_spam <- 0.30

P_spam_given_urgent <- (P_urgent_given_spam * P_spam) / P_urgent
P_spam_given_urgent
```

### Visualizing “Prior → Posterior”

```{r urgent-prior-posterior-plot, fig.height=4.5, fig.width=9}
df_pp <- tibble(
  belief = c("Prior: P(spam)", "Posterior: P(spam | Urgent)"),
  prob = c(P_spam, P_spam_given_urgent)
)

ggplot(df_pp, aes(x = belief, y = prob, fill = belief)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = percent(prob, accuracy = 0.1)),
            vjust = -0.5, fontface = "bold") +
  scale_y_continuous(labels = percent, limits = c(0, max(df_pp$prob) + 0.1)) +
  scale_fill_manual(values = c("#BBDEFB", "#FFCDD2"), guide = "none") +
  labs(
    title = "Bayes Update: How Evidence Changes Our Belief",
    subtitle = "A single word can dramatically increase the probability of spam",
    x = NULL, y = "Probability"
  ) +
  theme_minimal()
```

> **Beginner intuition:** The prior is your “base rate.” The posterior is your “base rate after seeing evidence.”

---

# Part 2: From Bayes to Naive Bayes

## 2.1 Classification Goal

In classification, we want to assign a class \(c\) (e.g., spam/ham) to a feature vector \(\mathbf{x}\) (e.g., words, counts, measurements).

Bayes says:

$$P(c \mid \mathbf{x}) = \frac{P(\mathbf{x} \mid c)\,P(c)}{P(\mathbf{x})}$$

To classify, we choose the class with the biggest posterior:

$$\hat{c}=\arg\max_c P(c\mid \mathbf{x})$$

## 2.2 The “Naive” Assumption (Conditional Independence)

If \(\mathbf{x} = (x_1, x_2, \ldots, x_p)\), Naive Bayes assumes:

$$P(\mathbf{x}\mid c) = \prod_{j=1}^p P(x_j \mid c)$$

In plain English:

> “Once I know the class, the features behave independently.”

Is that always true in real life? No.

So why does NB still work often?

- It can still rank classes correctly even if the probabilities are not perfect.
- Many problems (especially **text**) are “close enough” to independence for prediction.
- It’s a strong baseline that is hard to beat with small data.

## 2.3 Why We Use Logs (Underflow Problem)

Multiplying many small probabilities can become **numerically zero** in a computer.
So we compute:

$$\log P(c\mid \mathbf{x}) \propto \log P(c) + \sum_{j=1}^p \log P(x_j \mid c)$$

Same winner class, safer math.

---

# Part 3: Gaussian Naive Bayes (Numeric Features) — Iris Example

Gaussian NB is used when features are continuous (like lengths/widths). It assumes:

$$x_j \mid c \sim \mathcal{N}(\mu_{c,j},\sigma_{c,j}^2)$$

## 3.1 Data: Iris

```{r iris-data}
data(iris)
glimpse(iris)
```

### Visual overview (pair plot feel, but beginner-friendly)

```{r iris-pairs, fig.height=6.5, fig.width=10}
iris_long <- iris %>%
  pivot_longer(cols = -Species, names_to = "feature", values_to = "value")

ggplot(iris_long, aes(x = value, fill = Species)) +
  geom_histogram(bins = 22, alpha = 0.55, position = "identity") +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  scale_fill_manual(values = c("#1976D2", "#2E7D32", "#E53935")) +
  labs(
    title = "Iris Features by Species (Histograms)",
    subtitle = "NB will learn one distribution per feature per class",
    x = NULL, y = "Count", fill = "Species"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 3.2 Train/Test Split (70/30)

```{r iris-split}
set.seed(MASTER_SEED)
idx <- sample(1:nrow(iris), size = floor(0.7 * nrow(iris)))
train <- iris[idx, ]
test  <- iris[-idx, ]
```

## 3.3 Fit Naive Bayes (Gaussian)

We use `naivebayes::naive_bayes()` (as shown in the Lecture 07 slides).

```{r iris-fit}
nb_iris <- naive_bayes(Species ~ ., data = train)
nb_iris
```

## 3.4 Evaluate: Confusion Matrix + Accuracy

```{r iris-eval}
pred <- predict(nb_iris, test)
cm <- table(Predicted = pred, Actual = test$Species)
cm

acc <- sum(diag(cm)) / sum(cm)
acc
```

### Confusion matrix heatmap

```{r iris-cm-heatmap, fig.height=4.5, fig.width=7}
cm_df <- as.data.frame(cm)
ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = Freq), fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#4E79A7") +
  labs(
    title = "Confusion Matrix (Iris) — Naive Bayes",
    subtitle = sprintf("Test accuracy: %.1f%%", 100 * acc),
    x = "Actual class", y = "Predicted class", fill = "Count"
  ) +
  theme_minimal()
```

## 3.5 What Did NB Learn? (Means and Variances)

For Gaussian NB, each class learns a mean and variance per feature.

```{r iris-params}
nb_iris$tables
```

### Visualize class-conditional Gaussians (one feature)

We’ll look at `Petal.Length` because it separates species well.

```{r iris-gaussian-plot, fig.height=4.8, fig.width=9}
feat <- "Petal.Length"

params <- nb_iris$tables[[feat]] %>%
  as.data.frame() %>%
  rownames_to_column("Species") %>%
  rename(mean = V1, sd = V2)

x_grid <- seq(min(iris[[feat]]) - 0.5, max(iris[[feat]]) + 0.5, length.out = 400)

dens_df <- params %>%
  group_by(Species) %>%
  do(tibble(x = x_grid,
            density = dnorm(x_grid, mean = .$mean, sd = .$sd))) %>%
  ungroup()

ggplot() +
  geom_histogram(data = iris, aes(x = .data[[feat]], y = after_stat(density), fill = Species),
                 bins = 25, alpha = 0.25, color = "white") +
  geom_line(data = dens_df, aes(x = x, y = density, color = Species), linewidth = 1.2) +
  scale_fill_manual(values = c("#1976D2", "#2E7D32", "#E53935")) +
  scale_color_manual(values = c("#1976D2", "#2E7D32", "#E53935")) +
  labs(
    title = "Gaussian NB Assumption Visualized (Petal.Length)",
    subtitle = "Lines = learned normal distributions per class; bars = actual data histogram",
    x = feat, y = "Density", color = "Species", fill = "Species"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 3.6 Independence Assumption: Do Iris Features Look Independent?

They are not perfectly independent. Let’s visualize correlations.

```{r iris-corr-heatmap, fig.height=4.5, fig.width=7}
corr <- cor(iris[, 1:4])
corr_df <- as.data.frame(as.table(corr)) %>%
  rename(Feature1 = Var1, Feature2 = Var2, Correlation = Freq)

ggplot(corr_df, aes(x = Feature1, y = Feature2, fill = Correlation)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = round(Correlation, 2)), fontface = "bold", size = 3) +
  scale_fill_gradient2(low = "#E53935", mid = "white", high = "#1976D2",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(
    title = "Feature Correlations (Iris)",
    subtitle = "NB assumes conditional independence, but real features can be correlated",
    x = NULL, y = NULL, fill = "r"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

> **Key point:** Even with correlation, NB can still classify well.

---

# Part 4: Multinomial Naive Bayes (Counts) — Tiny Text Example

Multinomial NB is a common choice for text because text becomes **counts of words**.

## 4.1 A Tiny “Spam vs Ham” Dataset

This is intentionally small so you can see everything.

```{r tiny-text-data}
emails <- tibble(
  text = c(
    "urgent win money now",
    "urgent claim your prize",
    "win prize money money",
    "meeting schedule for tomorrow",
    "please review the report",
    "lunch meeting tomorrow"
  ),
  label = factor(c("spam", "spam", "spam", "ham", "ham", "ham"))
)
emails
```

## 4.2 Turn Text into Word Counts (Bag-of-Words)

```{r bow}
bow <- emails %>%
  mutate(id = row_number()) %>%
  unnest_tokens(word, text) %>%
  count(id, label, word, name = "n") %>%
  ungroup()

bow
```

### Vocabulary and document-term matrix

```{r dtm}
vocab <- sort(unique(bow$word))
dtm <- bow %>%
  tidyr::pivot_wider(names_from = word, values_from = n, values_fill = 0) %>%
  arrange(id)

dtm
```

## 4.3 Fit Multinomial NB

`naivebayes::naive_bayes()` automatically chooses an appropriate model type for predictors.
For word-count features, it uses a multinomial-like likelihood.

```{r multinom-fit}
train_x <- dtm %>% select(-id, -label)
train_y <- dtm$label

nb_text <- naive_bayes(x = train_x, y = train_y, laplace = 1)  # Laplace smoothing ON
nb_text
```

## 4.4 Predict a New Message

```{r text-predict}
new_msg <- tibble(text = "urgent money prize")

new_bow <- new_msg %>%
  mutate(id = 1) %>%
  unnest_tokens(word, text) %>%
  count(id, word, name = "n") %>%
  pivot_wider(names_from = word, values_from = n, values_fill = 0)

# Align columns to training vocab (add missing words with 0, drop extras)
for (w in setdiff(colnames(train_x), colnames(new_bow))) new_bow[[w]] <- 0
new_bow <- new_bow %>% select(all_of(colnames(train_x)))

pred_label <- predict(nb_text, new_bow)
pred_prob  <- predict(nb_text, new_bow, type = "prob")

pred_label
pred_prob
```

### Visualize predicted probabilities

```{r text-prob-plot, fig.height=4, fig.width=7}
prob_df <- as.data.frame(pred_prob) %>%
  pivot_longer(everything(), names_to = "class", values_to = "prob")

ggplot(prob_df, aes(x = class, y = prob, fill = class)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = percent(prob, accuracy = 0.1)),
            vjust = -0.5, fontface = "bold") +
  scale_y_continuous(labels = percent, limits = c(0, 1.05)) +
  scale_fill_manual(values = c(spam = "#E53935", ham = "#1976D2"), guide = "none") +
  labs(
    title = "Naive Bayes Prediction for a New Text Message",
    subtitle = "NB returns probabilities; we pick the larger one as the class",
    x = NULL, y = "Probability"
  ) +
  theme_minimal()
```

---

# Part 5: The Zero-Frequency Problem (and Laplace Smoothing)

## 5.1 The Problem

If a word never appeared in training for a class, NB would assign:

$$P(\text{word} \mid \text{class}) = 0$$

Then the whole product becomes zero:

$$\prod_j P(x_j \mid c) = 0$$

Even if the rest of the message screams “spam.”

## 5.2 Laplace Smoothing (Plain English)

Laplace smoothing adds a small amount to every count, so nothing is exactly zero.

In many textbooks this is described as:

- add 1 to each word count (**Laplace**)
- or add \(\alpha\) (**Lidstone**) where \(\alpha\) can be smaller than 1

We already used `laplace = 1` above.

### Quick demonstration with an “unseen word”

```{r smoothing-demo}
emails2 <- tibble(
  text = c("win money", "win prize", "team meeting", "project meeting"),
  label = factor(c("spam", "spam", "ham", "ham"))
)

bow2 <- emails2 %>%
  mutate(id = row_number()) %>%
  unnest_tokens(word, text) %>%
  count(id, label, word, name = "n") %>%
  pivot_wider(names_from = word, values_from = n, values_fill = 0)

x2 <- bow2 %>% select(-id, -label)
y2 <- bow2$label

nb_no_smooth <- naive_bayes(x2, y2, laplace = 0)
nb_smooth    <- naive_bayes(x2, y2, laplace = 1)

msg <- tibble(text = "win lottery")  # 'lottery' never seen in training
msg_bow <- msg %>%
  mutate(id = 1) %>%
  unnest_tokens(word, text) %>%
  count(id, word, name = "n") %>%
  pivot_wider(names_from = word, values_from = n, values_fill = 0)

for (w in setdiff(colnames(x2), colnames(msg_bow))) msg_bow[[w]] <- 0
msg_bow <- msg_bow %>% select(all_of(colnames(x2)))

prob_no  <- predict(nb_no_smooth, msg_bow, type = "prob")
prob_yes <- predict(nb_smooth,    msg_bow, type = "prob")

list(no_smoothing = prob_no, laplace_1 = prob_yes)
```

> **Key lesson:** In real NLP, smoothing is usually not optional — it prevents “single unseen word kills everything.”

---

# Part 6: Practical Tips (When NB Works Well)

```{r nb-pros-cons, echo=FALSE}
tbl <- tibble(
  Topic = c(
    "Speed",
    "Data size needs",
    "Good for text?",
    "Probability calibration",
    "Independence assumption",
    "Mixed feature types"
  ),
  Plain_English = c(
    "Very fast to train and predict (mostly counting + simple math).",
    "Often works well even with smaller datasets compared to more complex models.",
    "Yes — a classic baseline for spam filtering and document classification.",
    "Probabilities can be overconfident; use them carefully.",
    "Usually not true, but NB can still classify well.",
    "Works, but choose the right variant (Gaussian vs Multinomial vs Bernoulli)."
  )
)
knitr::kable(tbl, caption = "Naive Bayes: Pros & Cons (Beginner Summary)")
```

---

# Knowledge Check

<details>
<summary>**Question 1:** In your own words, what is a *prior* and what is a *posterior*?</summary>
**Answer:** The **prior** is what you believe *before* seeing new evidence (e.g., 20% of emails are spam). The **posterior** is what you believe *after* seeing evidence (e.g., given the word “Urgent”, spam might jump to 60%).
</details>

<br>

<details>
<summary>**Question 2:** Why is Naive Bayes called “naive”?</summary>
**Answer:** Because it assumes the features are conditionally independent given the class. In real life, features can be correlated, but NB pretends they’re independent to make the math simple and fast.
</details>

<br>

<details>
<summary>**Question 3:** What is the “zero-frequency problem” and how does Laplace smoothing fix it?</summary>
**Answer:** If a feature value never appeared in a class during training, NB can assign probability 0, which makes the whole class likelihood 0. Laplace smoothing adds a small amount to counts so probabilities never become exactly zero.
</details>

<br>

<details>
<summary>**Question 4:** When would you prefer Gaussian NB vs Multinomial NB?</summary>
**Answer:** **Gaussian NB** is for continuous numeric features (like lengths, temperatures). **Multinomial NB** is for counts (like word counts in documents).
</details>

---

# References

[1] Lecture 07 course material: `ANLY530-Machine-Learning/Week06/Lecture07-NB.pdf`

[2] Wikipedia contributors, “Naive Bayes classifier,” *Wikipedia, The Free Encyclopedia*, 2026. Available: `https://en.wikipedia.org/wiki/Naive_Bayes_classifier`

[3] scikit-learn developers, “Naive Bayes,” *scikit-learn User Guide*, 2026. Available: `https://scikit-learn.org/stable/modules/naive_bayes.html`

